Change in Mobility and Site Occupation during the Late Pleistocene in Korea

### 2019-11-22 analysis/paper/paper.Rmd
First draft is done, except for Umap part that Ben will cover!

# Discussion (still working on, needs more citations)

@@

3. Expedient vs curated
To monitor moibility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated). One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle]. Therefore, the curated technology indicates sites with special or higly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance. The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time].

During the transitoin period, sites were used in

# Conclusion (still working on)

This research aims to understand how the appearance of new technology affected hunter-gatherers in Korea and answer the three questions, how did the introduction of the new tools change the way people used the landscape, the way people used sites, and the population. To answer the question, we use HBE as a theoretical backgrounds and analyze stone artifacts from 22 sites dated before and after 40-35ka. 
The new technology is represented the stemmed points and other blade industry made out of new raw materials such as shale and rhyolite. We could clearly see that hunter-gatherers chose these materials particularly to make the newly added stone artifacts to the toolkit. Besides of technological change, we found that there are more varieties in the type of sites or groups during the transition period, both curated and expedient and then it becomes more expedient. We expected to see any pattern that would explain the relationship between environmental change and the emergence of the new technology. However, there was not enough environmental information as well as data to apply to verify the model. Therefore we focus on Radiocarbon dates as a proxy of population change based on the premise that population driven pressure could trigger people to invent a new tool. The results show that population dynamics might have had more impact on changes in stone artifact technology. ----

>>>>

# Discussion

@@

3. Expedient vs curated
To monitor mobility strategies and site occupation pattern, we examined the artifact density and retouch frequency, based on the premise that highly mobile foragers would have light weighted, highly retouched, replaceable, almost completely made tools and leave small amount of artifacts when they move to the new location (curated). One the other hand, less mobile foragers who tend to stay in one place yield more debitage and debris, produce less efficient tools since they can easily get supplied (expedient) [@clark2019landscapes; @kuhn1994formal; @kuhn2014mousterian; @meignen2006middle]. Therefore, the curated technology indicates sites with special or highly mobile groups with stone artifacts that are thoroughly designed and based on a high level of investment of time and effort in artifact manufacture and maintenance. The expedient technology represents base camp or residence with artifacts that tend to be manufactured, used, and abandoned in the short period of time [@vaquero2018searching;@binford1979organization; @torrence1989time]. There appear to be more variety in the type of sites (or group) during the transition period, both expedient or curated. It became more expedient at the end of the transition, which means foragers became less mobile and tended to stay in the base camp. Combined with the result of population dynamics, we can assume that the increased population made each group bigger and let them stay in one place.

# Conclusion

This research aims to understand how the appearance of new technology affected hunter-gatherers in Korea and answer the three questions, how did the introduction of the new tools change the way people used the landscape, the way people used sites, and the population. To answer the question, we use HBE as a theoretical backgrounds and analyze stone artifacts from 22 sites dated before and after 40-35ka. The new technology is represented the stemmed points and other blade industry made out of new raw materials such as shale and rhyolite. We could clearly see that hunter-gatherers chose these materials particularly to make the newly added stone artifacts to the toolkit. The results also show that the new technology enables the occupation of marginal environments at higher elevations. Besides of technological change, we found that there are more varieties in the type of sites or groups during the transition period, both curated and expedient and then it becomes more expedient. In other words, the foragers became less mobile and prone to stay in one place after the transition ended. We expected to see any pattern that would explain the relationship between environmental change and the emergence of the new technology. However, there was not enough paleoenvironmental information as well as data to apply to verify the model. Therefore we focus on Radiocarbon dates as a proxy of population change based on the premise that population driven pressure could trigger people to invent a new tool. The results show that population dynamics might have had more impact on changes in stone artifact technology.
